DOpElib: A Goal Oriented Software Library for Computing PDEs and Optimization Problems

### 2012-10-22 Paper_May_2012/main.tex
Added comment on linear solver for newton alg, and explanation for PDEProblemContainer

The evaluation of the residual will require the calculation of the integrals 
involved in the weak formulation of the problem. For example 
the $i$-th component of the residual, with respect to the given FE-basis 
$\phi_{h,i}$, will read as follows
\begin{align}\label{eq:residual_vector}
 a(u_h)(\phi_{h,i}) - f(\phi_{h,i}) = \sum_{K\in \mathcal T_h} \nu\int_K \nabla u, \nabla \phi_i\,dx + \ldots
 \end{align}
Again, we see that, except for the local element integrals, all 
calculations are independent from the concrete form of the PDE. 
Hence we define an object \texttt{Integrator}
that offers a method to calculate the residual given a user-defined 
object \texttt{pde} that implements the local element integrals, i.e.,
\begin{lstlisting}
  template<typename PROBLEM,typename VECTOR>
      void Integrator
             ::ComputeNonlinearResidual(PROBLEM& pde, 
                                        VECTOR & residual);
\end{lstlisting}
where \texttt{VECTOR} is one of the possible vector types given by \deal{}.
It is clear that we need to ensure that the integrator object has access
to the information stored in the \texttt{SpaceTimeHandler}
to correctly loop over all elements.
To solve the linearized systems, again, all that is left is the calculation
of the matrix corresponding to the linearized equation. To calculate this,
again we can define a method in the integrator that does exactly this provided
we have a user defined object \texttt{pde} that can evaluate the local 
integrals needed for assembling the matrix, i.e.,
\begin{lstlisting}
  template<typename PROBLEM,typename MATRIX>
      void Integrator
             ::ComputeMatrix(PROBLEM& pde, 
                             MATRIX & system_matrix);
\end{lstlisting}
where now \texttt{MATRIX} is one of the matrix types supported by \deal{}.
With this we can now define a nonlinear solver \texttt{NewtonSolver} 
for the solution of the discrete problem requiring the existence of the 
following function
\begin{lstlisting}
  template<typename PROBLEM, typename VECTOR>
      void NewtonSolver
             ::NonlinearSolve(PROBLEM& pde, 
                              VECTOR & solution);
\end{lstlisting}

@@

@@

\subsubsection{Nonstationary problems}
In this section, the 
numerical solution of nonstationary processes
is considered. We focus on details which are 
beyond stationary examples as discussed in the 
previous sections. Before, we introduce 
the steps of implementation in DOpElib, we 
derive the basic theoretical concepts.
As an example, we consider the heat equation:
Find $u\in V$:
\[
\partial_t u - \Delta u = 0
\]
and in weak formulation:
Find $u\in V$:
\[
a(u)(\phi) = 0 \quad \forall \phi \in V,
\]
with some appropriate space $V$ and 
\[
a(u)(\phi) = (\partial_t u, \phi) + (\nabla u, \nabla \phi) \quad \forall \phi \in V.
\]
Temporal discretization using the One-step-$\theta$ scheme reads:
Given the old timestep solution $u^n$, find $u^{n+1}$:
\[
\frac{(u^{n+1} - u^{n}, \phi)}{k} + \theta (\nabla u^{n+1}, \nabla \phi)
+ (1 - \theta) (\nabla u^{n}, \nabla \phi) = 0,
\]
Reordering in new and old terms and multiplication by the time step size $k$:
\[
(u^{n+1},\phi) + k \theta (\nabla u^{n+1}, \nabla \phi)
= (u^n, \phi) - (1 - \theta) (\nabla u^{n}, \nabla \phi).
\]
The concrete choice of $\theta$ leads to a
specific time stepping scheme. For instance,
$\theta = 0.5$ gives the Crank-Nicolson scheme,
\begin{lstlisting}
   CrankNicolsonProblem<...>
\end{lstlisting}
The main work is treated as for stationary problems
within 
\begin{lstlisting}
   MethodOfLines_SpaceTimeHandler<...>
\end{lstlisting}
Again, the key issue is that this procedure is totally independent 
of the specific form of the PDE. The user-related work 
remains to provide the time derivative $\partial_t u$ in
the implementation:
\begin{lstlisting}
  void CellTimeEquation(...)
   {
     // implement (u^{n+1},\phi) 
   }
\end{lstlisting}
and it is done in a straight-forward way:
\begin{lstlisting}
  for (all quadrature points per cell q)
   {
     Tensor<1,2> u = ...;

@@

\subsection{Solving of optimization problems}
Upgrading the framework for solving optimization problems
is described in this section. A prototypical problem reads:
\begin{align*}
\min\;&J(q,u) \\
  &\text{s.t.}\; a(q,u)(\phi) = 0 \quad \forall \phi\in V,\\
  &a \le q \le b,\\
  &g(q,u) \le 0,  
\end{align*}
where $u$ is a FE-function and $q$ can either be a FE-function or some 
fixed number of parameters, $a$ and $b$ are constraint bounds for the control $q$,
and $g(\cdot)$ is some state constraint.
As before, the main part is done in the \texttt{main} function there is no
great difference to the previous explications. However, 
since we utilize gradient-based optimization routines in the philosophy of
\cite{Troe99,HiKu01},
all derivatives must be implemented. All this information
must given to the \texttt{localpde.h} file. Herein, one has to 
implement the adjoint, tangent and adjoint Hession equations. 
The algorithmic parts are described in the literature \cite{BeMeVe06}.

>>>>

The evaluation of the residual will require the calculation of the integrals 
involved in the weak formulation of the problem. For example 
the $i$-th component of the residual, with respect to the given FE-basis 
$\phi_{h,i}$, will read as follows
\begin{align}\label{eq:residual_vector}
 a(u_h)(\phi_{h,i}) - f(\phi_{h,i}) = \sum_{K\in \mathcal T_h} \nu\int_K \nabla u, \nabla \phi_i\,dx + \ldots
 \end{align}
Again, we see that, except for the local element integrals, all 
calculations are independent from the concrete form of the PDE. 
Hence we define an object \texttt{Integrator}
that offers a method to calculate the residual given a user-defined 
object \texttt{pde} that implements the local element integrals, i.e.,
\begin{lstlisting}
  template<typename PROBLEM,typename VECTOR>
      void Integrator
             ::ComputeNonlinearResidual(PROBLEM& pde, 
                                        VECTOR & residual);
\end{lstlisting}
where \texttt{VECTOR} is one of the possible vector types given by \deal{}.
It is clear that we need to ensure that the integrator object has access
to the information stored in the \texttt{SpaceTimeHandler}
to correctly loop over all elements.
To solve the linearized systems, again, all that is left is the calculation
of the matrix corresponding to the linearized equation. To calculate this,
again we can define a method in the integrator that does exactly this provided
we have a user defined object \texttt{pde} that can evaluate the local 
integrals needed for assembling the matrix, i.e.,
\begin{lstlisting}
  template<typename PROBLEM,typename MATRIX>
      void Integrator
             ::ComputeMatrix(PROBLEM& pde, 
                             MATRIX & system_matrix);
\end{lstlisting}
where now \texttt{MATRIX} is one of the matrix types supported by \deal{}.
With this we can now define a nonlinear solver \texttt{NewtonSolver} 
for the solution of the discrete problem requiring the existence of the 
following function
\begin{lstlisting}
 template<typename LINSOLVE>
  template<typename PROBLEM, typename VECTOR>
      void NewtonSolver<LINSOLVE>
             ::NonlinearSolve(PROBLEM& pde, 
                              VECTOR & solution);
\end{lstlisting}
Where \texttt{LINSOLVE} denotes the solver that should be used to solve the 
linearized equations during the newton iteration.

@@

Finally, we will need to have an object that is capable of providing all 
the necessary functionality required from the template argument 
\texttt{PROBLEM} in all of the above. We provide a wrapper named
\texttt{PDEProblemContainer} collecting all the problem specific information 
like the description of the PDE and the functional, the boundary values, etc. 
Although, having such an object instead of directly giving these informations
to the \texttt{StatPDEProblem} seems like a nuicance at the moment, we will 
see that it is of greater interest for time-dependent problems or 
optimization problems in what follows in Section~\ref{sec:timedep} and
Section~\ref{sec:opt} respectively.

@@

\subsubsection{Nonstationary problems}\label{sec:timedep}
In this section, the 
numerical solution of nonstationary processes
is considered. We focus on details which are 
beyond stationary examples as discussed in the 
previous sections. Before, we introduce 
the steps of implementation in DOpElib, we 
derive the basic theoretical concepts.
As an example, we consider the heat equation:
Find $u\in V$:
\[
\partial_t u - \Delta u = 0
\]
and in weak formulation:
Find $u\in V$:
\[
a(u)(\phi) = 0 \quad \forall \phi \in V,
\]
with some appropriate space $V$ and 
\[
a(u)(\phi) = (\partial_t u, \phi) + (\nabla u, \nabla \phi) \quad \forall \phi \in V.
\]
Temporal discretization using the One-step-$\theta$ scheme reads:
Given the old timestep solution $u^n$, find $u^{n+1}$:
\[
\frac{(u^{n+1} - u^{n}, \phi)}{k} + \theta (\nabla u^{n+1}, \nabla \phi)
+ (1 - \theta) (\nabla u^{n}, \nabla \phi) = 0,
\]
Reordering in new and old terms and multiplication by the time step size $k$:
\[
(u^{n+1},\phi) + k \theta (\nabla u^{n+1}, \nabla \phi)
= (u^n, \phi) - (1 - \theta) (\nabla u^{n}, \nabla \phi).
\]
The concrete choice of $\theta$ leads to a
specific time stepping scheme. For instance,
$\theta = 0.5$ gives the Crank-Nicolson scheme,
\begin{lstlisting}
   CrankNicolsonProblem<...>
\end{lstlisting}
The main work is treated as for stationary problems
within 
\begin{lstlisting}
   MethodOfLines_SpaceTimeHandler<...>
\end{lstlisting}
Again, the key issue is that this procedure is totally independent 
of the specific form of the PDE. The user-related work 
remains to provide the time derivative $\partial_t u$ in
the implementation:
\begin{lstlisting}
  void CellTimeEquation(...)
   {
     // implement (u^{n+1},\phi) 
   }
\end{lstlisting}
and it is done in a straight-forward way:
\begin{lstlisting}
  for (all quadrature points per cell q)
   {
     Tensor<1,2> u = ...;

@@

\subsection{Solving of optimization problems}\label{sec:opt}
Upgrading the framework for solving optimization problems
is described in this section. A prototypical problem reads:
\begin{align*}
\min\;&J(q,u) \\
  &\text{s.t.}\; a(q,u)(\phi) = 0 \quad \forall \phi\in V,\\
  &a \le q \le b,\\
  &g(q,u) \le 0,  
\end{align*}
where $u$ is a FE-function and $q$ can either be a FE-function or some 
fixed number of parameters, $a$ and $b$ are constraint bounds for the control $q$,
and $g(\cdot)$ is some state constraint.
As before, the main part is done in the \texttt{main} function there is no
great difference to the previous explications. However, 
since we utilize gradient-based optimization routines in the philosophy of
\cite{Troe99,HiKu01},
all derivatives must be implemented. All this information
must given to the \texttt{localpde.h} file. Herein, one has to 
implement the adjoint, tangent and adjoint Hession equations. 
The algorithmic parts are described in the literature \cite{BeMeVe06}.
